Safe Redis Queries

### 2016-06-12 aplas16/sections/IndexedMonads.tex
Halfway editing the section in indexed monads.

Redis commands are \emph{actions}.
 We could capture the effects caused by an action, by expressing the states it
 affects, before and after. That is, the \emph{preconditions} and
 \emph{postconditions} of an action. In such way, we could also impose
 constraints on the preconditions.

\emph{Indexed monads} (or \emph{monadish},
 \emph{parameterised monad})\cite{indexedmonad}
 can be used\cite{typefun,staticresources} to model such preconditions and
 postconditions in types. An indexed monad is a type constructor that takes
 three arguments: an initial state, a final state, and the type of a value that
 an action computes, which can be read like a Hoare triple\cite{kleisli}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Class \ensuremath{\Conid{IMonad}} comes with two operations: \ensuremath{\Varid{unit}} for identities and \ensuremath{\Varid{bind}} for
compositions, as in monads.

We define a new datatype \ensuremath{\Conid{Popcorn}}, which is basically just
 the context \ensuremath{\Conid{Redis}} indexed with more information in types.
 We make it an instance of \ensuremath{\Conid{IMonad}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{9}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}{}\<[E]%
\\
\>[5]{}\hsindent{4}{}\<[9]%
\>[9]{}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The first and second argument of type \ensuremath{\Conid{Popcorn}} is where the
dictionaries going to be.

To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to
 \ensuremath{\Varid{unPopcorn}} to get an ordinary Hedis program back,
 with type information erased.

\subsection{\text{PING}: the first attempt}

In Redis, \text{PING} does nothing but replies with
 \text{PONG} if the connection is alive. In Hedis,
 \ensuremath{\Varid{ping}} has type:

>>>>

Stateful computations are often reasoned in a Hoare-logic style: each command
is labelled by a \emph{precondition} and a \emph{postcondition}. If the former
is satisfied before the command is executed, the latter is guaranteed to hold
afterwards.

In Haskell, stateful computations are represented by monads. In order to
reason about their behaviors within the type system, we wish to label a state
monad with its pre and postcondition. An \emph{indexed
monad}~\cite{indexedmonad} (also called \emph{monadish} or
\emph{parameterised monad}) is a monad that, in addition to the type of value
it computes, takes two more type arguments representing an initial state and
a final state, to be interpreted like a Hoare triple~\cite{kleisli}:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{class}\;\Conid{IMonad}\;\Varid{m}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{p}\;\Varid{p}\;\Varid{a}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}\to (\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b})\to \Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The intention is that a computation of type \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} is a stateful computation
such that if it starts execution in a state satisfying \ensuremath{\Varid{p}} and terminates, it
yields a value of type \ensuremath{\Varid{a}}, and the new state satisfies \ensuremath{\Varid{q}}.
The operator \ensuremath{\Varid{unit}} lifts a pure computation to a stateful computation that
does not alter the state. In \ensuremath{\Varid{x}\mathbin{`\Varid{bind}`}\Varid{f}}, a computation \ensuremath{\Varid{x}\mathbin{::}\Varid{m}\;\Varid{p}\;\Varid{q}\;\Varid{a}} must
be chained before \ensuremath{\Varid{f}\mathbin{::}\Varid{a}\to \Varid{m}\;\Varid{q}\;\Varid{r}\;\Varid{b}}, which expects a value of type \ensuremath{\Varid{a}} and
a state satisfying \ensuremath{\Varid{q}} and, if terminates, ends in a state satisfying \ensuremath{\Varid{r}}.
The result is a monad \ensuremath{\Varid{m}\;\Varid{p}\;\Varid{r}\;\Varid{b}} --- a computation that, if executed in a state
satisfying \ensuremath{\Varid{p}} and terminates, yields a value \ensuremath{\Varid{b}} and a state satisfying \ensuremath{\Varid{r}}.
Indexed monads have been used ~\cite{typefun,staticresources} ... \todo{for what? Some discriptions here to properly cite them.}

We define a new indexed monad \ensuremath{\Conid{Popcorn}} which, at term level, merely wraps
\ensuremath{\Conid{Redis}} in an additional constructor. The purpose is to add the pre and
postconditions at type level:
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{newtype}\;\Conid{Popcorn}\;\Varid{p}\;\Varid{q}\;\Varid{a}\mathrel{=}\Conid{Popcorn}\;\{\mskip1.5mu \Varid{unPopcorn}\mathbin{::}\Conid{Redis}\;\Varid{a}\mskip1.5mu\}~~,{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{instance}\;\Conid{IMonad}\;\Conid{Popcorn}\;\mathbf{where}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{unit}\mathrel{=}\Conid{Popcorn}\mathbin{\cdot}\Varid{return}{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\Varid{bind}\;\Varid{m}\;\Varid{f}\mathrel{=}\Conid{Popcorn}\;(\Varid{unPopcorn}\;\Varid{m}\bind \Varid{unPopcorn}\mathbin{\cdot}\Varid{f})~~.{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
To execute a \ensuremath{\Conid{Popcorn}} program, simply apply it to \ensuremath{\Varid{unPopcorn}} to erase the additional type information and get back an ordinary \Hedis{} program.

\paragraph{\text{PING}: A First Example}
In \Redis{}, \text{PING} does nothing but replies with
 \text{PONG} if the connection is alive. In Hedis,
 \ensuremath{\Varid{ping}} has type:
